### 2020-07-13 manuscript/paper_BRM.Rmd
shorten results sections

author: 
  - name          : "Malte Valentin Lüken"

authornote: |
  Affiliation: University of Amsterdam, Graduate School of Psychology, Department of Methodology and Statistics.

  This manuscript constitutes the internship report as part of the Research Master's Psychology.

  Student number: 12750166. Email address: malte.luken@student.uva.nl.

  Specialization: Methodology and Statistics. Supervisor: dhr. dr. Ingmar Visser.

  I would like to thank Simon Kucharsky and Ingmar Visser for their helpful advice on developing and implementing this research project.

@@

bibliography      : "report.bib"
appendix          : 
 - "report_appendix.Rmd"
 - "report_technical_appendix.Rmd"

@@

For how long did humans seek to open a window into each other's minds, revealing their hidden cognitive processes? While many methods aiming to achieve this have been developed in the last decades, eye-tracking as one of them has become more and more reliable, accurate, and widely applied in cognitive psychology [@Duchowski2017]. Eye-tracking is typically used to study cognitive processes involving attention and information search based on recorded gaze position [@Schulte-Mecklenbeck2017]. Before these processes can be studied, the raw gaze data is classified into events that are distinct in their physiological patterns (e.g., duration), underlying neurological mechanisms, or cognitive functions [@Leigh2015]. Commonly distinguished events are fixations, saccades, smooth pursuit, and post-saccadic oscillations (PSOs). Classifying eye-tracking data reduces their complexity and is the first step towards cognitive interpretation [@Andersson2017; @Salvucci2000].  
The classification is normally done by algorithms, which is considered faster, more objective, and reproducible compared to human coding [@Andersson2017]. @Hein2017 give a comprehensive overview of different classification algorithms [for a structured review on classifying saccades, see also @Stuart2019]. 
As illustrated in Figure \@ref(fig:plot-algorithm-workflow), most classification algorithms consist of four steps: First, the raw gaze position is transformed into a metric of position and/or time. Second, a filtering or smoothing procedure is applied to the data to separate the gaze signal from noise and artifacts [@Spakov2012]. Third, depending on the method and settings of the algorithm, each sample is labeled as a candidate for one of the predefined events. Fourth, the algorithm decides which candidates to accept, relabel, or merge [@Hessels2017; @Komogortsev2010].  
A frequently implemented method uses probabilistic models to classify events. Instead of discrete classification, these models assign probabilities for belonging to an event to each sample. Probabilistic models have the advantage that they learn parameters from the data, can adapt to the task- and individual-specific gaze signals, and can be easily applied online [@Kasneci2014, p. 324].
One class of probabilistic models that are used in eye movement classification are hidden Markov models (HMMs). Figure \@ref(fig:plot-hmm-structure) illustrates the structure of HMMs: Parallel to the gaze signal evolves a sequence of distinct states that cannot be directly observed. Each gaze sample depends on its corresponding state. Each state depends on the previous but not on earlier states of the sequence [@Cappe2005; @Visser2019].  
In the context of eye movements, HMMs are suitable probabilistic models because the hidden states can be interpreted as events and gaze data are dependent time series. On this basis, several classification algorithms using HMMs have been developed.  
One instance is described in @Salvucci2000 and combines the HMM with a threshold approach (named "Identification by HMM" [I-HMM]). Samples are first labeled as fixations or saccades, depending on whether their velocity exceeds a threshold, and then reclassified by the HMM. Recently, @Pekkanen2017 developed an algorithm that filters the position of gaze samples through naive segmented linear regression (NSLR). The algorithm uses an HMM to parse the resulting segments into fixations, saccades, smooth pursuits, and PSOs based on their velocity and angle (named NSLR-HMM). Another version by @Mihali2017 uses a Bayesian HMM to separate microsaccades (short saccades during fixations) from motor noise based on sample velocity (named "Bayesian Microsaccade Detection" [BMD]). Moreover, @Houpt2018 developed a hierarchical approach that describes sample velocity and acceleration through an autoregression (AR) model, computes the regression weights through an HMM, and estimates the number of events with a beta-process (BP) from the data (named BP-AR-HMM).  
Several studies have tested the performance of HMM algorithms against other classification methods: I-HMM has been deemed as robust against noise, behaviorally accurate, and showing a high sample-to-sample agreement with a human coder [@Andersson2017; @Komogortsev2010; @Salvucci2000]. However, the agreement was lower when compared to an algorithm using a Bayesian mixture model [@Kasneci2014; @Tafaj2012]. NSLR-HMM showed even higher agreement to human coding than I-HMM but was outperformed by a recently developed algorithm using convolutional neural networks [@Bellet2019; @Pekkanen2017].
In sum, HMMs seem to be a promising method for classifying eye movements. Still, the existing HMM algorithms each have their weaknesses that could be improved upon (and that could explain their inferior performance in comparison to other methods).  
First, I-HMM relies on setting an appropriate threshold, which can distort the results [@Blignaut2009; @Komogortsev2010; @Shic2008]. Second, the current implementation of NSLR-HMM requires human-coded data, which limits its applicability. It also inherits fixed parameters that prevent the algorithm to adapt to the individual- or task-specific signals. Third, BMD limits the classification to microsaccades which are irrelevant in many applications and sometimes even considered as noise [@Duchowski2017]. The opposite problem was observed for BP-AR-HMM: It tends to estimate an unreasonable number of events from the data of which many are considered as noise events. Therefore, the authors suggest using it as an exploratory tool followed by further event classification [@Houpt2018].  
HMMs can also be used as a generative model to make inferences about the underlying processes of eye movements, a property that has been rarely used in the context of classification [cf. @Mihali2017]. For example, an HMM with four events could be compared to one with three events to see whether an event is present in the data. 
I propose to develop an algorithm that uses HMMs to classify eye movement data. It will address the weaknesses of previous HMM algorithms and utilize the generative property of HMMs.

@@

(ref:plot-algorithm-workflow) Example workflow for eye movement event classification algorithms: (a) the raw gaze signal is transformed into a velocity signal (in deg/s); (b) the velocity signal is smoothed and filtered; (c) samples are labeled as events (indicated by colors), and (d) relabeled. Sequences of samples belonging to the same event are merged (indicated by black segments).

```{r plot-algorithm-workflow, fig.cap="(ref:plot-algorithm-workflow)", fig.height = 8}

set.seed(126)

velRaw <- sample(c(rgamma(25, 3, 1/6), rgamma(7, 25, 1/5), rgamma(8, 15, 1/5)), 40)

sample <- 1:length(velRaw)

velSmooth <- as.numeric(smooth(velRaw, kind = "3R"))

eventPre <- ifelse(velSmooth > 40, "s", "f")

eventPost <- ifelse(velSmooth < 60, "f", eventPre)

df <- data.frame(sample, velRaw, velSmooth, eventPre, eventPost)

# Create plot

p1 <- ggplot(df, aes(sample, velRaw)) + geom_point()

p2 <- ggplot(df, aes(sample, velSmooth)) + geom_point()

p3 <- ggplot(df, aes(sample, velSmooth)) + geom_point(aes(color = eventPre))

p4 <- ggplot(df, aes(sample, velSmooth)) + geom_point(aes(color = eventPost))

plotList <- list(p1, p2, p3, p4)

for(p in 1:length(plotList)) {

  plotList[[p]] <- plotList[[p]] + scale_y_continuous(name = "Velocity\n(deg/s)", 
                                                      limits = c(0, 200), breaks = c(0, 100, 200)) + 
    xlab("") + geom_line(color = "grey") + ggtitle(letters[p]) + 
    theme(legend.position = "none")

plotList[[4]] <- plotList[[4]] + xlab("Sample") +
  geom_segment(aes(x = 1, y = 180, xend = 7, yend = 180), size = 1.2) +
  geom_segment(aes(x = 8, y = 180, xend = 16, yend = 180), size = 1.2) +
  geom_segment(aes(x = 17, y = 180, xend = 40, yend = 180), size = 1.2)

grid.arrange(plotList[[1]], plotList[[2]], plotList[[3]], plotList[[4]], ncol = 1)

(ref:plot-hmm-structure) Example structure of HMMs: The observed velocity signal *V* depends on hidden states *S* (displayed for six samples). Each state depends on the previous state. Colors indicate different states. Figure inspired by Visser and Speekenbrink [-@Visser2019, p. 133].

```{r plot-hmm-structure, fig.cap="(ref:plot-hmm-structure)", fig.height=2, fig.width=4}

knitr::include_graphics(here("manuscript/internship_report_files/external-files/hmm-structure.png"), auto_pdf = T, dpi = 400)

## Research Question
Can HMMs improve describing and classifying eye-tracking data? This is the general research question of my proposal and can be specified as following: Are HMMs useful generative models for recorded eye movements that improve classifying them into events? This research question will guide the development of the algorithm and leads to two hypotheses: First, I predict that HMMs are useful generative models for eye movements. This hypothesis will be supported if the algorithm yields events that match theoretical expectations for benchmark data. Other results will be treated as evidence against the hypothesis. Second, I hypothesize that HMMs will improve the classification performance of eye movements. I will treat it as support for this hypothesis if the algorithm outperforms algorithms using other methods when applied to benchmark data. I will treat other results as evidence against the hypothesis.

# Operationalization
The proposed algorithm aims to improve the drawbacks of previous HMM algorithms. Therefore, its classification method is oriented to four major goals: (a) it does not require parameter settings through the user (e.g., thresholds) or (b) human-labeled data as input; (c) it covers the most relevant eye movement events, namely fixations, saccades, smooth pursuit, and PSOs; (d) it confirms rather than explores the presence of events in the data. Besides implementing the algorithm in the statistical programming language R [@RCoreTeam2020], I will develop its concrete technical design exploratorily.  
The first step will be choosing an appropriate metric that allows separating the gaze data into events. @Zemblys2018 provide an overview of different gaze data metrics and compare how well they predict human-labeled events. The next step will choosing a filtering/smoothing method to reduce the noise in the gaze data [for a comparison of different gaze data filters, see @Spakov2012]. The criteria for choosing a metric and a filter will be its accuracy and not relying on user parameter settings.  
The generative part of the algorithm will allow users to build an HMM which matches their theoretical assumptions about the gaze data. The algorithm will take a preprocessed gaze data metric (e.g., filtered velocity) as input and optimize the model parameters. The output of the model will be probabilities for hidden states. The states can be interpreted as event labels for each sample. For instance, the model could assign to a sample a probability of 0.6 for belonging to a fixation and 0.4 for belonging to a saccade. These probabilities can be used to compute the hidden state sequence (e.g., by choosing events with the highest probability). Moreover, the model will output the parameters of the distribution of labeled samples for each event. It will allow generating samples from these distributions to simulate gaze data.  
Before testing the hypotheses, I will conduct a simulation study to estimate the parameter recovery of the generative model: Artificial eye movement data will be generated by the model. The parameters of the model will be varied across sensible ranges. Then, the model will be applied to the data. Finally, the estimated results of the model will be compared to the true parameters that have been used to generate the data.  
The first hypothesis will be tested by comparing different generative models on benchmark data. The models will differ in their assumed number of events in the data. The model comparison will result in a preferred model which scores best on the evaluation criterion. The preferred model is expected to match the theoretically assumed number of present events in the benchmark data.  
To test the second hypothesis, the proposed algorithm will be compared against other algorithms that have also been applied to benchmark data. The performance of the proposed algorithm will be assessed by metrics that are commonly used in classification (e.g., Cohen's Kappa).

## Sample Characteristics
The hypotheses will be tested on two benchmark data sets that have been published in previous studies. The first set was published in @Andersson2017 and contains eye-tracking data from 17 subjects. Three types of stimuli were presented on a screen ($1024 \times 768$ pixels) to each subject: A static image, a linearly moving dot, and a video. Subjects were instructed to look freely at the images and follow the movements of the dots and the objects in the videos. Their eye movements were recorded with a Hi-Speed 1250 tower-mounted eye-tracker (SensoMotoric Instruments) and a sampling rate of 500 Hz. Only data from the right eye were used. The data were labeled by two human experts in eye movement research as fixations, saccades, smooth pursuits, PSOs, or blinks [@Andersson2017, p. 621].
The second data set, published by @Ehinger2019, consists of eye-tracking data from 15 subjects who performed six blocks of 10 different experimental tasks. Here, I will only consider data from Tasks 4 and 5 because they are qualitatively different from the first data set. In Task 4, subjects were instructed to fixate a central target on the screen ($1920 \times 1080$ pixels) for 20 s. In Task 5, subjects also fixated a central target but were instructed to blink each time they heard a beep. Seven beeps with a duration of 100 ms and a 1.5 s interval in between were presented. For both tasks, eye-movements were recorded by two eye-trackers simultaneously (EyeLink 1000, SR Research; and Pupil Labs glasses, Pupil Labs) with a sampling rate of 250-500 Hz and 240 Hz (interpolated), respectively [@Ehinger2019, pp. 3-11].

## Data Analysis
I will treat the parameter recovery of the generative model as satisfactory when it yields a 95% CI covering the true parameters. The median rate of correct classifications should be a minimum of 90%. 
The first hypothesis will be tested using a model comparison approach: For both data sets, models with one, two, three, four, and five eye movement events plus a noise event will be compared. The noise event will represent blinks. For the first data set and static images, I expect the model with three events (excl. noise) to be the preferred model since smooth pursuit movements are not assumed to be present. For moving dots and video stimuli, I expect the model with four events (excl. noise) to be preferred, since smooth pursuits are expected in the data.
For the second data set, I expect a model with one event (excl. noise) to be preferred because only fixations should be present in the data from both tasks. For all comparisons, the preferred model will be chosen according to the highest Schwarz weight [@Wagenmakers2004].
I will test the second hypothesis using the metrics that have been used in the original study of the first data set: @Andersson2017 compared ten algorithms to human-labeled data by the root mean squared error (RMSE) of event distribution descriptives, the sample-to-sample agreement indicated by Cohen's Kappa, and the disagreement ratio across all samples indicated by the confusion matrix. I expect the algorithm with a generative model assuming four events (excl. noise) to achieve either a lower RMSE, higher Cohen's Kappa, or lower disagreement ratio than all the other algorithms which are evaluated in the study.

## Intended Results
Preferred generative models that match the theoretical assumptions on the number of events in the data would support the notion that HMMs describe gaze data appropriately. If a generative model with fewer events than expected is preferred, this would mean that the model is not able to distinguish the events well enough. If more events than expected are preferred, this would indicate that the model is describing noise or an unexpected event is present in the data.
Outperforming other algorithms in agreement to human-coded data would demonstrate that the proposed algorithm classifies events similar to human coders. However, @Hooge2018a observed considerable differences in how humans code gaze data. They concluded that human coders cannot be seen as a gold standard for event classification. 
The proposed algorithm would be the first allowing the user to conduct model comparisons as part of the classification. This comparison could be used to investigate the data more thoroughly on a qualitative (e.g., assumed present events) or quantitative (e.g., event sample distributions) level.
In contrast to many other classification algorithms, the proposed algorithm would not only classify gaze data but also explain why it is classified this way by using a generative model. This outstanding property of the algorithm would guarantee its competitiveness against other popular algorithms, especially the recently developed machine learning approaches which do not make generative assumptions [e.g., @Bellet2019]. It could spark a shift from pure classification output towards understanding eye movements on a more fundamental level.

@@

# Algorithm Development
As part of answering my research questions, I developed an algorithm named gazeHMM to classify gaze data into discrete eye movement events. The following section describes the development of gazeHMM and its final version that has been used to obtain results. Technical details can be found in Appendix B.

## Eye Movement Metrics
Many different metrics can be used to describe gaze data and eye movement events [@Zemblys2018]. Here, my goal was to find those metrics that separate the gaze data samples belonging to different events the best. However, many metrics rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007; @Zemblys2018]. This can be problematic because such parameters are often set without theoretical justification and they differ substantially between metrics. Because they belong to the most basic metrics which do not require parameter settings, I used velocity, acceleration, and sample-to-sample angle [synonymous to relative or change in angle; @Larsson2013] in gazeHMM.  
Theoretically, these three metrics should separate eye movement events clearly. Fixations typically inherit samples with low velocity and acceleration [@Larsson2013]. Due to tremor, the angle between samples should not follow any direction but a random walk [@Duchowski2017]. In contrast, saccade samples usually have a high velocity and acceleration and follow the same direction. PSO samples tend to have moderate velocity and high acceleration since they occur between saccades and low velocity events [@Larsson2013]. They can be specifically distinguished by their change in direction clustered around 180 degrees [@Pekkanen2017]. Importantly, the oscillations depend on the resolution of the gaze recording: Eye-trackers with higher sampling frequency yield more changes in direction and more samples in between those changes. Those samples in between typically follow the same direction. Thus, with high sampling frequencies, PSO samples might also cluster around a sample-to-sample angle of zero with outliers around 180 degrees. Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015]. Other algorithms focus exclusively on classifying microsaccades [e.g., @Mihali2017], but as proposed earlier, they are not in the scope of gazeHMM.

## Filtering and Smoothing
As with metrics, many methods can be used to filter or smooth gaze data before, while, or after computing eye movement metrics. Their purpose is to remove noise or artifacts from the gaze data that could distort the classification [@Duchowski2017]. Filtering (smoothing) methods that are applied before computing the eye movement metrics target the recorded gaze position. While removing noise, filters might also erase differences in the metrics between samples. For instance, tremor movements during fixations could be filtered out and in consequence, the samples would follow the same direction. This could complicate separating fixations from smooth pursuits. The same problem applies to PSOs and saccades where filtering could erase oscillations (see Figure \@ref(fig:filter-angle)).

(ref:filter-angle) Example data from @Andersson2017 displayed as sample-to-sample angle (in rad) over time (in s). Line colors indicate whether the positional data has been filtered before computing the sample-to-sample angle (red) or not (blue). For this example, I used a Butterworth filter of order three and a normalized cutoff frequency of 0.3. The full line segment marks where potential oscillations are filtered out. The dashed line segment illustrates where potentially random change in angle is filtered out.

```{r filter-angle, fig.cap="(ref:filter-angle)"}

filter.ex <- A2017$video[[3]] %>%
  mutate(x = px2va(x, dim[1], res[1], dist),
         y = px2va(y, dim[2], res[2], dist),
         angle = calc_theta(x, y),
         x.filt = as.numeric(signal::filter(signal::butter(3, 0.3), x)),
         y.filt = as.numeric(signal::filter(signal::butter(3, 0.3), y)),
         angle.filt = calc_theta(x.filt, y.filt),
         t = (t-t[1])/1e6) %>%
  dplyr::filter(t > 0.5, t < 0.75)

ggplot(filter.ex) + 
  geom_path(aes(x = t, y = force_neg_pi_pi(angle)), color = "blue") +
  geom_path(aes(x = t, y = force_neg_pi_pi(angle.filt)), color = "red") +
  annotate(geom = "segment", x = 0.525, xend = 0.54, y = -3, yend = -3, color = "black", size = 0.8) +
  annotate(geom = "segment", x = 0.635, xend = 0.67, y = -3, yend = -3, color = "black", size = 0.8, linetype = 6) +
  scale_y_circular(name = "Sample-to-sample angle (in rad)", units = "radians", limits = c(-pi, pi)) +
  scale_x_continuous(name = "Time (in s)")

Instead of filtering before computing the metrics, I decided to combine both in one step. Previous algorithms have used two methods that both filter and compute derivatives of the gaze position (i.e., velocity and acceleration). @Houpt2018 computed the first and second discrete derivative (similar to a Sobel and Laplace filter, respectively) of the gaze position to be used by their algorithm. Similarly, @Nystrom2010 used a Savitzky-Golay (SG) filter [@Savitzky1964] to estimate velocity and acceleration from a gaze signal. Figure \@ref(fig:filter-comparison) displays an example of velocity and acceleration signals obtained by both methods. They follow a similar trend, but the SG velocity signal was less noisy and the acceleration signal was consequently lower than the Laplace acceleration signal. Therefore, I chose to implement the SG filter to compute velocity and acceleration signals, because it filters out more noise than the discrete derivatives but still preserves the edges in the signal to distinguish between events. To preserve motor noise in the sample-to-sample angle signal, gazeHMM computes the absolute angle as the backward difference and the first discrete derivative as the forward difference (see Appendix B).

(ref:filter-comparison) Example data from @Andersson2017 displayed as velocity (deg/s) and acceleration (in deg/s$^2$) over time (in s). Line colors indicate which filter has been used to derive the signal. Savitzky-Golay filters were applied with order three and length five (corresponding to 10 ms).

```{r filter-comparison, fig.cap="(ref:filter-comparison)"}

filter.comp <- A2017$video[[3]] %>%
  mutate(x = px2va(x, dim[1], res[1], dist),
         y = px2va(y, dim[2], res[2], dist),
         vel.x.sg = signal::sgolayfilt(x, m = 1),
         vel.y.sg = signal::sgolayfilt(y, m = 1),
         vel.sg = sqrt(vel.x.sg^2 + vel.y.sg^2)*fr,
         acc.x.sg = signal::sgolayfilt(x, m = 2),
         acc.y.sg = signal::sgolayfilt(y, m = 2),
         acc.sg = sqrt(acc.x.sg^2 + acc.y.sg^2)*fr^2,
         vel.x.sb = lead(x) - lag(x),
         vel.y.sb = lead(y) - lag(y),
         vel.sb = sqrt(vel.x.sb^2 + vel.y.sb^2)*fr,
         acc.x.lp = lead(x) - 2*x + lag(x),
         acc.y.lp = lead(y) - 2*y + lag(y),
         acc.lp = sqrt(acc.x.lp^2 + acc.y.lp^2)*fr^2,
         t = (t-t[1])/1e6) %>%
  pivot_longer(c("vel.sg", "vel.sb", "acc.sg", "acc.lp"), names_to = c("metric", "filter"), names_sep = "[:punct:]", ) %>%
  mutate_at(c("metric", "filter"), as.factor) %>%
  dplyr::filter(t > 0.5, t < 1.5)

filter.comp$metric <- factor(filter.comp$metric, labels = c("Acceleration", "Velocity"))

ggplot(filter.comp, aes(x = t, y = value, color = filter)) + 
  geom_path() + facet_wrap(vars(metric), nrow = 2, scales = "free_y", strip.position = "left") + 
  scale_x_continuous(name = "Time (in s)") +
  scale_y_continuous(name = "") +
  scale_color_discrete(name = "Filter", labels = c("Laplace", "Sobel", "Savitzky-Golay")) +
  theme(strip.placement = "outside")

## The Generative Model
The generative model underlying gazeHMM is a multivariate HMM. It can contain between two and four states that correspond to different eye movement events: The first state always represents fixations, the second saccades, the third PSOs, and the fourth smooth pursuits. Thus, users can choose whether they would like to classify only fixations and saccades, or additionally PSOs and/or smooth pursuits.  
In general, HMMs consist of three submodels: An initial state model, a transition model, and a response model. The initial state model contains probabilites for the first state of the hidden sequence. The evolution of the sequence is in turn described by the trantision model, which comprises the probabilities for transitioning between different states in the HMM. The response model encompasses distributions describing the response variables for every state in the model.  
In gazeHMM, the response model has three response variables which are the velocity and acceleration signals obtained by the SG filter and the change in angle signal. The response variables are treated as conditionally independent on the states. Conditional independence might not accurately resemble the relationship between velocity and acceleration (which are naturally correlated). This step was merely taken to keep the HMM simple and identifiable.  
Previous algorithms using HMMs have used Gaussian distributions to describe velocity and acceleration signals (sometimes after log-transforming them). However, several reasons speak against choosing the Gaussian: First, both signals are usually positive (depending on the computation). Second, the distributions of both signals appear to be positively skewed conditionally on the states and third, to have variances increasing with their mean. Thus, instead of using the Gaussian, it could be more appropriate to describe velocity and acceleration with a distribution that follows these three properties. In gazeHMM, I use gamma distributions with a shape and scale parametrization for this purpose. It has to be noted that the gamma was chosen out of convenience and the best fitting distribution might be different between eye-trackers, subjects, and tasks.   
To model the sample-to-sample angle, I pursued a novel approach in gazeHMM: Using a mixture of von Mises distributions (with a mean and concentration parameter) and a uniform distribution. Both the distributions and the metric operate on the full unit circle (i.e., between 0 and $2\pi$), which should lead to rather symmetric distributions. Because the fixations change their direction similar to a random walk, their sample-to-sample angle can be modeled by a uniform distribution. Thus, the uniform distribution should distinguish fixations from the other events.  
The initial state model and the transition model use multinomial distributions (with a category for each state). For all three submodels of the HMM, no time-varying covariates are included. Since the implementation of gazeHMM allows for regression models for each submodel and state, this implies that only intercepts for each parameter are estimated.

## Missing Data and Blinks
HMMs can estimate parameters and hidden states despite missing data. They either treat data as missing at random or as missing depending on states [@Visser2019]. In gazeHMM, I assume data to be missing at random. However, most of the missing data in eye movement classification are due to blinks. Therefore, the user can indicate in gazeHMM which samples should be labeled as blinks (other missing samples are treated as noise). Often, eye-trackers record a few samples with unreasonably high velocity and acceleration before losing the pupil signal when a blink occurs. Since these samples could distort the classification of saccades in the HMM, I decided to remove them heuristically. Before classifying the samples, gazeHMM sets all samples within 50 ms before and after blink samples as missing. The window of 50 ms was rather motivated empirically than theoretically and can bet set to any value the user considers appropriate.  
When evaluating the likelihood of the HMM, given the parameters and joint density of response distributions, for missing data, the model integrates over all possible values. The resulting density for missing samples is therefore set to one [@Visser2019].

## Optimization and Classification
The parameters of the HMM are estimated through maximum likelihood using an expectation-maximization (EM) algorithm [@Dempster1977; @McLachlan1997]. The EM algorithm is generally suitable to estimate likelihoods with missing variables. For HMMs, it imputes missing with expected values and iteratively maximizes the joint likelihood of parameters conditional on the observed data (velocity, acceleration, and sample-to-sample angle) and the expected hidden states [eye movement events; @Visser2019]. The sequence of hidden states is estimated through the Viterbi algorithm [@Viterbi1967; @ForneyJr1973] by maximizing the posterior state probability. Parameters of the response distributions (except for the uniform distribution) are optimized on the log-scale (except for the mean parameter of the von Mises distribution) using a spectral projected gradient method [@Birgin2000] and Barzilai-Borwein step lengths [@Barzilai1988].

## Postprocessing
After classifying gaze samples into states, gazeHMM applies a postprocessing routine to the estimated state sequence. I implemented this routine because constraining the transition probabilities for PSOs to turn into non-saccade events to zero often caused PSOs not to appear in the state sequence at all. Moreover, gazeHMM does not explicitly control the duration of events in the HMM which occasionally led to unreasonably short events. Thus, the postprocessing routine heuristically compensates for such violations. This routine relabels one-sample fixations and smooth pursuits, saccades with a duration below a minimum threshold, and PSOs that follow non-saccade events. Samples are relabeled as the state of the previous event. Finally, samples initially indicated as missing are labeled as noise (including blinks) and event metrics are computed (e.g., fixation duration).

## Implementation
The algorithm is implemented in R [version: 3.6.3; @RCoreTeam2020] and uses the packages signal [@Ligges2015] to compute velocity and acceleration signals, depmixS4 [@Visser2010] for the HMM, and BB [@Varadhan2009] for Barzilai-Borwein spectral projected gradient optimization. The algorithm is available on GitHub (www.github.com/maltelueken/gazeHMM).

# Simulation Study
To assess how well the HMM recovers parameters and state sequences, I conducted a simulation study. The design and analysis of the study were preregistered on the Open Science Framework (https://doi.org/10.17605/OSF.IO/VDJGP). The major part of this section is a direct copy of the preregistration. When appropriate, additional explanations were added and tenses adapted to make the design easier to read and understand.  
In the study, the HMM repeatedly generated data with a set of parameters (true parameter values). The same model was then applied to estimate the parameters from the generated data (estimated parameter values). I compared the true with the estimated parameter values to assess whether a parameter was recovered by the model. Additionally, I compared the true states of the HMM with the estimated states to judge how accurately the model recovered the states that generated the data.

## Starting Values
The HMM always started with a uniform distribution to estimate the initial state and state transition probabilities. To generate random starting values for the estimation of shape, scale, and concentration parameters, I used gamma distributions with a shape parameter of $\alpha_{start}=3$ and $\beta_{start}=\psi_{true}/2$ with $\psi_{true}$ being the true value of the parameter to be estimated. This setup ensured that the starting values were positive, their distributions were moderately skewed, and the modes of their distributions equaled the true parameter values. Mean parameters of the von Mises distribution always started at their true values.

>>>>

author: 
  - name          : "Malte Lueken"
    affiliation   : "1"
    corresponding : yes
    address       : "Postal address"
    email         : "malte_lueken@arcor.de"
  - name          : "Simon Kucharsky"
    affiliation   : "1"
  - name          : "Ingmar Visser"
    affiliation   : "1"

affiliation:
  - id            : "1"
    institution   : "Department of Psychology, University of Amsterdam. Amsterdam, The Netherlands"

authornote: |
  This is an author note.

@@

bibliography      : "report.bib"

@@

@@

@@

# Simulation Study
To assess how well the HMM recovers parameters and state sequences, we conducted a simulation study. The design and analysis of the study were preregistered on the Open Science Framework (https://doi.org/10.17605/OSF.IO/VDJGP). This section is majorly copied from the preregistration (with adapted tenses). The study was divided in four parts. Here, we only report the first two parts, the other two parts, which address starting values and missing data, can be found in the supplementary material (URL).  
The HMM repeatedly generated data with a set of parameters (true parameter values). The same model was applied to estimate the parameters from the generated data (estimated parameter values). We compared the true with the estimated parameter values to assess whether a parameter was recovered by the model. Additionally, we contrasted the true states of the HMM with the estimated states to judge how accurately the model recovered the states that generated the data.

## Starting Values
The HMM always started with a uniform distribution to estimate the initial state and state transition probabilities. Random starting values for the estimation of shape, scale, and concentration parameters were generated by gamma distributions with a shape parameter of $\alpha_{start}=3$ and $\beta_{start}=\psi_{true}/2$, with $\psi_{true}$ being the true value of the parameter to be estimated. This setup ensured that the starting values were positive, their distributions were moderately skewed, and the modes of their distributions equaled the true parameter values. Mean parameters of the von Mises distribution always started at their true values.

## Design
In the first part, we varied the parameters of the HMM. For models with $k \in \{2, 3, 4\}$ states, $q \in \{10, 15, 20\}$ parameters were manipulated, respectively. For each parameter, the HMM generated 100 data sets with $N = 2500$ samples and the parameter varied in a specified interval in equidistant steps. This resulted in $100 \times (10+15+20) = 4500$ recoveries. Only one parameter alternated at once, the other parameters were set to their default values. All parameters of the HMM were estimated freely (i.e., there were no fixed parameters in the model). We did not manipulate the initial state probabilities because these are usually irrelevant in the context of eye movement classification. For the transition probabilities, we only simultaneously changed the probabilities for staying in the same state (diagonals of the transition matrix) to reduce the complexity of the simulation. The left over probability mass was split evenly between the probabilities for switching to a different state (per row of the transition matrix). Moreover, we did not modify the mean parameters of the von Mises distributions: As location parameters, they do not alter the shape of the distribution and they are necessary features for the HMM to distinguish between different states.  
We defined approximate ranges for each response variable (see supplementary material) and chose true parameter intervals and default values so that they produced samples that roughly corresponded to these ranges. Tables \@ref(tab:sim-parameters-trans) and \@ref(tab:sim-parameters-resp) show the intervals and default values for each parameter in the simulation. Parameters were scaled down by factor 10 (compared to the reported ranges) to improve fitting of the gamma distributions. We set the intervals for shape parameters of the gamma distributions for all events to [1,5] to examine how skewness influenced the recovery (shape values above five approach a symmetric distribution). The scale parameters were set so that the respective distribution approximately matched the assumed ranges. Since the concentration parameters of the von Mises distribution are the inverse of standard deviations, they were varied on the inverse scale. An example of simulated data from the HMM with default parameters is visualized in Figure \@ref(fig:plot-sim-ex).  
In the second part, we manipulated the sample size of the generated data and the amount of noise added to it. The model parameters were set to their default values. For models with $k \in \{2, 3, 4\}$ states and sample sizes of $N \in \{500, 2500, 10000\}$, we generated 100 data sets ($100 \times 3 \times 3 = 900$ recoveries). These samples sizes roughly corresponded to small, medium, and large eye-tracking data sets for a single participant and trial. To simulate noise, we replaced velocity and acceleration values $y$ with draws from a gamma distribution with $\alpha_{noise} = 3$ and $\beta_{noise}=(y/2)\tau_{noise}$ with $\tau_{noise} \in [1,5]$ varying between data sets. This procedure ensured that velocity and acceleration values remained positive and were taken from moderately skewed distributions with modes equal to the original values. To angle, we added white noise from a von Mises distribution with $\mu_{noise} = 0$ and $\kappa_{noise} \in 1/[0.1,10]$ varying between data sets. $\tau_{noise}$ and $\kappa_{noise}$ were increased simultaneously in equidistant steps in their intervals.

## Data Analysis
For each parameter, we calculated the root median square proportion deviation [RMdSPD; analogous to root median square percentage errors, see @Hyndman2006] between the true and estimated parameter values: $$\text{RMdSPD} = \sqrt{\text{Med}\left((\frac{\psi_{true}-\psi_{est}}{\psi_{true}})^2\right)}.$$
Even though it was not explicitly mentioned in the preregistration, this measure is only appropriate when $\psi_{true} \ne 0$. This was not the case for some mean parameters of the von Mises distributions. In those cases, we used $\psi_{true} = 2\pi$ instead. We treated $\text{RMdSPD} < 0.1$ as good, $0.1 \le \text{RMdSPD} < 0.5$ as moderate, and $\text{RMdSPD} \ge 0.5$ as bad recovery of a parameter. By taking the median, we reduced the influence of potential outliers in the estimation and using proportions enabled us to compare RMdSPD values across parameters and data sets.  
Additionally, we applied a bivariate linear regression with the estimated parameter values as the dependent and the true parameter values as the independent variable to each parameter that has been varied on an interval in part one. Regression slopes closer to one indicated that the model better captured parameter change. Regression intercepts different from zero reflected a bias in parameter estimation.  
To assess state recovery, we computed Cohen's kappa (for all events taken together, not for each event separately) as a measure of agreement between true and estimated states for each generated data set. Cohen's kappa estimates the agreement between two classifiers accounting for the agreement due to chance. Higher kappa values were interpreted as better model accuracy. We adopted the ranges proposed by @Landis1977 to interpret kappa values.
Models that could not be fitted were excluded from the recovery.
